---
title: "Homework 2"
author: "Troy Layouni"
date: 2019-09-20
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The first step is to load the tidyverse package

```{r packages} 
library(tidyverse) 
```

## Question 1

Reading in the Captain Trash Wheel data from the first excel sheet titled "Mr. Trash Wheel" as well as:

  * Omitting non-data entries 
  * Cleaning the names of variables
  * Omitting rows not including dumpster-specific data
  * Rounding the "Sports Ball" variable to the nearest integer and converting it to an integer variable using as.integer

```{r, trash_wheel_Data}
library(readxl)

trash_wheel_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "Mr. Trash Wheel", 
  range = "A2:N408") %>% 
  janitor::clean_names() %>%
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
```

**Question 1, part 2: Reading in and cleaning precipitation data**

  * Omitting rows without data
  * Adding a year variable
  * Combining of both data on precipitation for 2017 and 2018
  * Converting the month number to a character variable for month name

Reading in and cleaning "2017 Precipitation" data

```{r precipitation_data_2017}
precipitation_2017_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2017 Precipitation",
    range = "A2:B14") %>% 
    janitor::clean_names() %>%
    mutate(year = 2017) 
```

Reading in and cleaning "2018 Precipitation" data

```{r precipitation_data_2018}
precipitation_2018_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2018 Precipitation",
    range = "A2:B14") %>% 
    janitor::clean_names() %>%
    mutate(year = 2018)    
```

Combining precipitation data for 2017 and 2018 and converting month into a character variable

```{r precipitation_data_full}
precipitation_data_full =
  bind_rows(precipitation_2017_data, precipitation_2018_data) %>% 
  mutate(month = month.name[month])
```

**Question 1, part 3: Describing the precipitation data**

The Captain Trash Wheel dataset, `trash_wheel_data` includes data on the contents of each dumpster filled from 2014-2019. There are `r ncol(precipitation_data_full)` columns in this dataset, with information on debris collected such as plastic bags and sports balls, with a median of `r median(pull(filter(trash_wheel_data, year == 2017), sports_balls))` balls in each dumpster in 2017, as well as information on dumpster volume and weight, which averaged  `r mean(pull(filter(trash_wheel_data, year == 2018), weight_tons))` tons per dumpster in 2018, `r mean(pull(filter(trash_wheel_data, year == 2017), weight_tons))` tons in 2017, `r mean(pull(filter(trash_wheel_data, year == 2016), weight_tons))` tons in 2016,  `r mean(pull(filter(trash_wheel_data, year == 2015), weight_tons))` tons in 2015, and `r mean(pull(filter(trash_wheel_data, year == 2014), weight_tons))` tons in 2014. The dataset also gives the number of homes that were powered by Captain Trash Wheel. Data on precipitation in Baltimore, where Captain Trash Wheel is located, was collected and combined into our dataset `precipitation_data_full`, which includes `r ncol(precipitation_data_full)` columns for the variables month, year and total precipitation, and `r nrow(precipitation_data_full)` rows, one for every month of 2017 and 2018. In 2017, Baltimore, received a total of `r sum(pull(precipitation_2017_data, total))` inches of precipitation and in 2018 received `r sum(pull(precipitation_2018_data, total))` inches. The mean monthly precipitation for 2017 was `r mean(pull(precipitation_2017_data, total))` inches (standard deviation of `r sd(pull(precipitation_2017_data, total))` inches) and for 2018 the mean was `r mean(pull(precipitation_2018_data, total))` inches (standard deviation of `r sd(pull(precipitation_2018_data, total))` inches). 

## Question 2

**Question 2, part 1: cleaning politician party affiliation data "pols-month" by:**

* breaking up data variable into month, day, and year 
* creating a new variable called "president"
* removing "day" variable
* replacing month number with month name 

```{r political_party_data}
political_party_data = 
read_csv(file = "./data/fivethirtyeight_datasets/pols-month.csv") %>%
  separate(mon, sep = "-", into = c("year", "month", "day")) %>% 
  mutate(president = if_else(prez_dem < 1, "gop", "dem")) %>% 
  select(-day, -prez_dem, -prez_gop) %>% 
  mutate(month = month.name[(as.integer(month))]) %>% 
  mutate(year = as.numeric(year))
```

**Question 2, part 2: cleaning the stock market index data by:**

* breaking up data variable into month, day and year 
* replacing month number with month name 
* removing "day" variable 
* organize data according to year and month
* arrange data so that year and month are the first columns

```{r stock_market_data}
stock_market_data = 
read_csv(file = "./data/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>% 
  separate(date, sep = "/", into = c("month", "day", "year")) %>% 
  mutate(month = as.factor(month)) %>% 
  arrange(year, month) %>%
  select(-day, year, month, close) %>% 
  mutate(month = month.name[(as.integer(month))]) %>% 
  mutate(year = as.numeric(year))
```

**Question 2, part 3: tidying the unemployment data by:** 

* use pivot_longer to create a month variable and percentage of unemployment variable 
* clean data 

```{r unemployment_data}
unemployment_data = 
read_csv(file = "./data/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names() %>% 
  pivot_longer(jan:dec, names_to = "month", values_to = "unemploy_percent") %>% 
  mutate(month = recode(month, "jan" = "January", "feb" = "February", "mar" = "March", "apr" = "April", "jun" = "June", "jul" = "July", "aug" = "August", "sep" = "September", "oct" = "October", "nov" = "November", "dec" = "December")) %>% 
  mutate(year = as.numeric(year))

# mutate(month = match(month, month.abb)) 
 # mutate(month = month.name[month]) 
```

**Question 2, part 4: Joining datasets by merging political_party_data and stock_market_data, then merging unemployment into the result**

Merging `stock_market_data` into `political_party_data`

```{r political_stock_data}
political_stock_data = 
  left_join(political_party_data, stock_market_data, by = c("year", "month"))
```

Merging `unemployment_data` into `political_stock_data`

```{r fivethirtyeight_data}
fivethirtyeight_data = 
  left_join(political_stock_data, unemployment_data, by = c("year", "month"))
```

**Question 2, part 5: Summary of FiveThirtyEight data**

The FiveThirtyEight dataset, `fivethirtyeight_data` contains information on political leaders party affiliation as well as stock market closing values and unemployment rates by year and month. Variables like `sen_dem` and `rep_gop` give the number of democratic senators and GOP representatives, respectively, and `president` tells us the party affiliation of the president of the United States (dem, or gop). This data contains `r ncol(fivethirtyeight_data)` rows, each representing different month and year in history, and `r nrow(fivethirtyeight_data)` columns. We can see that between `r min(pull(fivethirtyeight_data, year))` and `r max(pull(fivethirtyeight_data, year))`, unemployment (variable `unemploy_percent`) at its lowest was `r min(pull(fivethirtyeight_data, unemploy_percent), na.rm = TRUE)` % and at its highest was `r max(pull(fivethirtyeight_data, unemploy_percent), na.rm = TRUE)`%. We can also see that the mean closing value of the Standard & Poor (S&P) stock market index (variable `close`) over this time period was `r mean(pull(fivethirtyeight_data, close), na.rm = TRUE)`.

## Question 3

**Question 3, part 1: Loading and tidying Popular baby names data by:**

* removing duplicate rows 
* creating uniform case structure
* creating consistent formatting for the ethnicity variable

```{r baby_names_data}
baby_names_data = 
read_csv(file = "./data/Popular_Baby_Names.csv") %>%
  janitor::clean_names() %>% 
  mutate(childs_first_name = str_to_upper(childs_first_name)) %>%
  mutate(ethnicity = recode(ethnicity, "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER", "BLACK NON HISP" = "BLACK NON HISPANIC", "WHITE NON HISP" = "WHITE NON HISPANIC")) %>% 
  distinct()
```

**Question 3, part 2: creating a table for rank in popularity of the name Olivia for females over time**

```{r olivia_table}
baby_names_data %>%
  filter(childs_first_name == "OLIVIA", gender == "FEMALE") %>% 
  select(ethnicity, year_of_birth, rank) %>% 
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "rank"
  ) %>% 
  knitr::kable(caption = "Popularity Rank of the Name Olivia by Ethnicity from 2011-2016")
```

**Question 3, part 3: creating a table for most popular boy's name over time**

```{r boy_names_table}
baby_names_data %>%
  filter(rank == "1", gender == "MALE") %>% 
  select(ethnicity, year_of_birth, childs_first_name) %>% 
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "childs_first_name"
  ) %>% 
  knitr::kable(caption = "Most Popular Male Name by Ethnicity from 2011-2016")
```

**Question 3, part  4: creating a scatter plot of the number white male boys born in 2016 by rank of their respective names**

```{r rank_count_plot}
baby_names_data %>%
  filter(ethnicity == "WHITE NON HISPANIC", gender == "MALE", year_of_birth == "2016") %>% 
  ggplot(aes(x = rank, y = count)) +
  geom_point(color = "blue") +
  ggtitle("Number of Non-Hispanic White Males Born in 2016 with Ranked Names") + 
  xlab("Ranking of Name") + 
  ylab("Number of Children")
```








