---
title: "Homework 2"
author: "Troy Layouni"
date: 2019-09-20
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**The first step is to load the tidyverse package**

```{r packages} 
library(tidyverse) 
```

## Question 1

Reading in the Mr. Trash Wheel data from the first excel sheet titled "Mr. Trash Wheel" as well as:

  * Omitting non-data entries 
  * Cleaning the names of variables
  * Omitting rows not including dumpster-specific data
  * Rounding the "Sports Ball" variable to the nearest integer and converting it to an integer variable using as.integer

```{r, Trash_Wheel_Data}
library(readxl)

Trash_Wheel_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  col_names = TRUE, 
  sheet = "Mr. Trash Wheel", 
  skip = 1, 
  range = "A2:N408") %>% 
  janitor::clean_names() %>%
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(sports_balls))
```

**Question 1, part 2: Reading in and cleaning precipitation data**

  * Omitting rows without data
  * Adding a year variable
  * Combining of both data on precipitation for 2017 and 2018
  * Converting the month number to a character variable for month name

Reading in and cleaning 2017 Precipitation data

```{r Precipitation_Data_2017}
Precipitation_2017_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    col_names = TRUE,
    sheet = "2017 Precipitation",
    skip = 1,
    range = "A2:B14") %>% 
    janitor::clean_names() %>%
    mutate(year = 2017) 
```

Reading in and cleaning 2018 Precipitation data

```{r Precipitation_data_2018}
Precipitation_2018_data = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    col_names = TRUE,
    sheet = "2018 Precipitation",
    skip = 1,
    range = "A2:B14") %>% 
    janitor::clean_names() %>%
    mutate(year = 2018)    
```

Combining precipitation data for 2017 and 2018 and creating a character month variable 

```{r Precipitation_data_full}
Precipitation_data_full =
  bind_rows(Precipitation_2017_data, Precipitation_2018_data) %>% 
  mutate(month = month.name[month])
```

**Question 1, part 3: Describing the precipitation data**
Mr. Trash Wheel, 
The data collected for precipitation in 2017 has `r nrow(Precipitation_2017_data)` rows, one precipitation measure for each month, and has `r ncol(Precipitation_2017_data)` columns, one for month, total precipitation, and year. In 2017, Baltimore, where Mr. Trash Wheel is located received a total of `r sum(pull(Precipitation_2017_data, total))` inches of precipitation. The data collected for precipitation in 2018 also contains `r nrow(Precipitation_2018_data)` rows, and has `r ncol(Precipitation_2018_data)` columns, including the same information on total precipitation, month and year for 2018. According to these data, in 2018 Baltimore received `r sum(pull(Precipitation_2018_data, total))` inches of precipitation.
These data were combined into one data frame containing `r nrow(Precipitation_data_full)` rows, one for every month of 2017 and 2018 and `r ncol(Precipitation_data_full)` columns. 



The mean monthly precipitation for 2017 is `r mean(pull(Precipitation_2017_data, total))` inches (standard deviation of `r sd(pull(Precipitation_2017_data, total))` inches)

The mean monthly precipitation is `r mean(pull(Precipitation_2018_data, total))` inches (standard deviation of `r sd(pull(Precipitation_2018_data, total))` inches). 







